---
title: "R Notebook"
author: jakobpovsic
date: 07/06/2020
output: html_notebook
---
# Introduction

In this document we are exploring gun violence statistics in the United states.
Data was downloaded from https://www.kaggle.com/jameslko/gun-violence-data/data

## Data visualisations
```{r}
library(tidyverse)
library(ggplot2)

data <- read.csv("data/gun-violence-data_01-2013_03-2018.csv")
pop_data <- read.csv("data/datasets_28951_36892_State Populations.csv")

df1 <- data%>%
  group_by(state)%>%
  summarise(total_n_killed=sum(n_killed), total_n_injured=sum(n_injured), n_total = n())%>%
  rowwise()%>%
  mutate(im_ratio=total_n_killed/total_n_injured)%>%
  mutate(pop=pop_data[pop_data$State == state, ]$X2018.Population)%>%
  mutate(killed_per_capita=(total_n_killed / pop) * 100000, injured_per_capita = (total_n_injured / pop) * 100000, total_n_per_capita= (n_total / pop) * 100000)%>%
  arrange(desc(im_ratio))
```
### Number of reports per 100,000 per state
```{r}
ggplot(df1) +
  geom_col(aes(x = reorder(state, total_n_per_capita), y = total_n_per_capita)) +
  labs(y = 'Reports per 100,000', x = 'State') +
  coord_flip()
```
### Murders per capita per state
```{r}
ggplot(df1) +
  geom_col(aes(x = reorder(state, killed_per_capita), y = killed_per_capita)) +
  labs(y = 'Murders per 100,000', x = 'State') +
  coord_flip()
```
### Gun violence through time
```{r}
df4 <- data%>%
  group_by(date)%>%
  summarise(n_killed = sum(n_killed), n_injured = sum(n_injured), n_reports = n())

ggplot(df4, aes(x = date, group=1)) +
  geom_line(aes(y = n_killed, color='Murders')) +
  geom_line(aes(y = n_injured, color='Injuries')) +
  geom_line(aes(y = n_reports, color='Incidents'))

```

### Ratio of suicides to all gun related deaths
This is a really interesting statistic, according to data more than half of all gun related deaths are suicides.
I've marked a report as suicide if there was only a single person killed and no-one injured and the incident characteristics contained the word suicide.
```{r}
df2 <- data%>%
  mutate(suicide=grepl('suicide', incident_characteristics) & n_killed == 1 & n_injured == 0)%>%
  group_by(state)%>%
  summarise(n_suicides = sum(suicide), n_total = sum(n_killed))%>%
  mutate(total_suicide_ratio = n_suicides / n_total)

ggplot(df2) +
  geom_col(aes(x = reorder(state, total_suicide_ratio), y = total_suicide_ratio)) +
  labs(y = 'Ratio of suicides in all gun violence reports', x = 'State') +
  coord_flip()
```
```{r}
ggplot(df2) +
  geom_col(aes(x = reorder(state, n_total), y = n_total)) +
  geom_col(aes(x = reorder(state, n_total), y = n_suicides), fill='red') +
  labs(x = "State", y = "Total gun violence related deaths (suicides in red)") +
  coord_flip()
```
## Interpretation
Looking at these statistics it paints a very different picture of the US, where most gun deaths are a result of a person taking its own life.
This is very contradictory with the cultural perception of vast majority of gun deaths being a result of criminal activity which is in reality a minority of gun related deaths.

Looking at per capita reports there is an apparent trend of more crime in the south of the country.

## References
- https://www.kaggle.com/jameslko/gun-violence-data/data
